A Large-Scale Study of Polymorphism in R

### 2019-04-06 paper/main.tex
touched up discussion of per fun poly

\AT{Working through this now.}

Now we wonder: {\it what kinds of function-level annotations are useful?}

For instance, should we link argument types with return types?
Consider Figure~\ref{fig:funanno}.
\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
iden <- function(x) x;

@@

Harkening back to our discussion of character/real polymorphism, we mentioned that we might want to consider such a type signature to be a list or vector index only if a list or vector was present as another argument.
Now, this may be moving well beyond the purview of R programmers, but perhaps there is value in such a scenario, particularly if we include even {\it more} information in a type system.
For example, say we have a function which takes in a 2D point list.
There are a plethora of useful annotations, here, and of varying degrees of granularity.
We might want simply a "list" annotation, or we might want a "list<double>" annotation to be more specific.
In addition, though, we might want to know if the list has a names attribute, yielding "list<double>@names".
Then, perhaps it would be useful to know what the names were, so "list<double>@names\{x, y\}".

Now, depending on the granularity of individual argument annotations, we might want to link certain arguments together.
Consider the code snippet in Figure~\ref{fig:indexof}.
\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
access <- function(l, i) l[[i]]

L <- list(a=1, b=2, c=3)
access(L, 2) # => 2
access(L, "c") # => 3

# possible annotations:
# [1] l: list, i: {character, real} -> real
# [2] l: list<double>, i: {character, real} -> real
# [3] l: list<X>, i: {character, real} -> X
# [4] l: list<double>@names, i: index -> double
# [5] l: list<double>@names{a, b, c}, i: index@indexOf(l) -> double
\end{lstlisting}}\caption{Linking annotations.}\label{fig:indexof}\end{figure}
The annotations could be used to generate code to check for compliance, like a contract.

>>>>

We saw that argument-level signatures and polymorphism are insufficient to capture some interesting patterns at the function level.
For instance, consider Figure~\ref{fig:funanno}.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
iden <- function(x) x;

@@

In certain cases, language implementations could benefit from knowing the return type of a function when the function is fed certain arguments.
In Figure~\ref{fig:funanno}, we can assign the function a more sophisticated signature if we don't simply collapse the observed argument signatures element-wise.
This gets more interesting with attributes, as we might want to annotate a function to communicate that it adds an attribute to whatever the list of attributes of its input value.

One drawback of function-level polymorphism is that functions with many arguments tend to be sensitive to small changes in argument types, which inflates the number of unique signatures observed.
We saw this in Figure~\ref{fig:L2funcounts}, where the distribution of signatures per function was roughly in line with those from Figures~\ref{fig:L0funcounts} and~\ref{fig:L1funcounts} but the peak (i.e., the function with the most signatures) was far more pronounced.
That said, overall the number of polymorphic functions did not increase by too large a margin considering the added expressivity of {\bf L2}'s types.
